---
title: "Homework 4 - Working with large datasets"
author: "Eunchong Kang"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

Library setup
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(data.table)
library(microbenchmark)
library(tidyverse)
library(Matrix)
library(sparsesvd)
library(grid)
library(gridExtra)

```

# Question 1 - benchmark data.table's grouping on real data

0. Import data as data.table
```{r}
dt <- fread("../Scorecard_2009-2016.csv")
dt2 <- dt
```

1. Modify data types with the level

        | C | Value              |     |
        |---|--------------------|-----|
        | 1 | Public             | pub |
        | 2 | Private nonprofit  | pnp |
        | 3 | Private for-profit | pfp |
        
```{r warning=FALSE}
dt2$TUITFTE <- as.numeric(dt2$TUITFTE)
dt2$SAT_AVG <- as.numeric(dt2$SAT_AVG)
dt2$CONTROL <- as.factor(dt2$CONTROL)
levels(dt2$CONTROL) <- c("pub", "pnp", "pfp")
```

2. Tabulate the number of schools by `CONTROL` with or without non-NA values for both `TUITFTE` *and* `SAT_AVG`.

```{r}
# The number of schools by CONTROL
dt2[, .N, by = CONTROL]

# The numbet of schools by CONTROL that have non-NA values for TUITFTE and SAT_AVG
dt2[TUITFTE != "NA" & SAT_AVG != "NA", .N, by = CONTROL]
# or dt2[!is.na(TUITFTE) & !is.na(SAT_AVG), .N, by = CONTROL]
```

3. Compute the mean and SD of `TUITFTE` and `SAT_AVG` by `CONTROL`.

```{r}
# My NA-removed mean and sd function
mean_NA_removed <- function(x){
        return(mean(x[!is.na(x)]))
}
sd_NA_removed <- function(x){
        return(sd(x[!is.na(x)]))
}

# Using data.table
dt2[,
    .(TUITFTE_mean=mean_NA_removed(TUITFTE), SAT_AVG_mean=mean_NA_removed(SAT_AVG),
      TUITFTE_sd=sd_NA_removed(TUITFTE), SAT_AVG_sd=sd_NA_removed(SAT_AVG)), 
    by = CONTROL]

# Using aggregate
merge(aggregate(dt2[, c("TUITFTE", "SAT_AVG")], dt2[, "CONTROL"], mean_NA_removed),
      aggregate(dt2[, c("TUITFTE", "SAT_AVG")], dt2[, "CONTROL"], sd_NA_removed), 
      by = "CONTROL", suffixes = c("_mean", "_sd"), sort = FALSE)

# Compare two methods
microbenchmark(dt2[,
                   .(TUITFTE_mean=mean_NA_removed(TUITFTE), SAT_AVG_mean=mean_NA_removed(SAT_AVG),
                     TUITFTE_sd=sd_NA_removed(TUITFTE), SAT_AVG_sd=sd_NA_removed(SAT_AVG)), 
                   by = CONTROL],
               merge(aggregate(dt2[, c("TUITFTE", "SAT_AVG")], dt2[, "CONTROL"], mean_NA_removed),
                     aggregate(dt2[, c("TUITFTE", "SAT_AVG")], dt2[, "CONTROL"], sd_NA_removed), 
                     by = "CONTROL", suffixes = c("_mean", "_sd"), sort = FALSE),
               times = 10)
```


# Question 2- doing more with "by" in data.table

1. Make `scores.sub`
```{r}
# Making `scores.sub`
scores.sub <- na.omit(dt2, cols=c("TUITFTE", "SAT_AVG"))
```

2. Make a plot of `SAT_AVG` over `TUITFTE`
```{r}
# Making a plot
ggplot(scores.sub, aes(TUITFTE, SAT_AVG)) +
        geom_point(aes(color= CONTROL), alpha = 0.7, position = "jitter") +
        ggtitle("Scatter Plot of SAT_AVG over TUITFTE") +
        coord_cartesian(xlim=c(0,40000), ylim=c(500,1600)) +
        labs(x= "Tuition per FTE", y="Average SAT") +
        scale_x_continuous(breaks=seq(0, 40000, 5000)) +
        scale_y_continuous(breaks=seq(500, 1600, 100))
```

3. Tabulate the number of schools that have tuition per FTE over
20,000 and/or average SAT over 1200, grouped by ownership
category.

```{r}
scores.sub[, .N, keyby = .(CONTROL, 
                        TUITFTE_over20000 = TUITFTE > 20000, 
                        SAT_AVG_over1200 = SAT_AVG > 1200)]
```

# Question 3 - subsets of data 

1. Obtain the tuition per FTE and average SAT for the
two schools with the top average SAT within each ownership
group.

```{r}
scores.sub[, .(CONTROL, TUITFTE, SAT_AVG)][order(-SAT_AVG)][, head(.SD, 1), keyby = CONTROL]
```


# Question 4 - MovieLens sparse dataset

0. Importing data
```{r, message=FALSE}
links <- read_csv("../ml-latest-small/links.csv")
movies <- read_csv("../ml-latest-small/movies.csv")
ratings <- read_csv("../ml-latest-small/ratings.csv")
tags <- read_csv("../ml-latest-small/tags.csv")
```

1. How many of the movies have the "Comedy" genre attached to
them? 

```{r}
sum(grepl("Comedy", movies$genres))
```

2. Build a sparse matrix of the movies by users and investigate the range, quartiles,
etc. of number of movies rated by each user.

```{r}
# Making sparse matrix of movies by user.
mtrx <- sparseMatrix(ratings$movieId, ratings$userId, x=1)
mtrx_ratings <- mtrx[rowSums(mtrx) != 0,]
dim(mtrx_ratings)

# Summary
summary <- summary(mtrx_ratings) %>% 
        group_by(j) %>% 
        summarise(count = n())
ggplot(summary, aes(y=count)) +
                geom_boxplot(width=1) +
                scale_x_continuous(limits = c(-1, 1)) +
                scale_y_continuous(breaks=seq(0, 3000, 500)) +
                ggtitle("Boxplot of the number of ratings by users") +
                theme(axis.text.x=element_blank())

summary(summary$count)
```

3. Compute SVD

```{r}
# Making key variable to match the spare matrix row number to movieId 
keys <- data.frame(movieId = unique(ratings$movieId), key = 1:9724)
keys_byRow <- left_join(keys, movies, by="movieId")

# Computing SVD
svd <- sparsesvd(mtrx_ratings, rank=3)
dim(svd$u)
```

4. Plot the columns of the U matrix against each other: 1 vs 2, 2 vs 3, 1
vs 3.

```{r fig.width=9, fig.height=5}
# Change svd matrix to data.frame with key variable
svd_df <- data.frame(svd$u) %>%
        mutate(key = 1:nrow(svd$u)) %>%
        rename("u1"="X1", "u2"="X2", "u3"="X3")

plot_1_2 <- ggplot(svd_df, aes(u1, u2)) +
        geom_point() +
        ggtitle("Scatter plot of u2 over u1")

plot_1_3 <- ggplot(svd_df, aes(u1, u3)) +
        geom_point() +
        ggtitle("Scatter plot of u3 over u1")

plot_2_3 <- ggplot(svd_df, aes(u2, u3)) +
        geom_point() +
        ggtitle("Scatter plot of u3 over u2")

grid.arrange(plot_1_2, plot_1_3, plot_2_3, nrow = 1, ncol = 3)

```

5. Investigate the names of these movies. What property can you
infer about the top 6 movies in the tail w.r.t. column 1 and 3.

```{r}
# Top 6 moives w.r.t the sum of u1 and u3
df_movies_key_u <- left_join(keys_byRow, svd_df, by="key")
df_movies_key_u %>% mutate(sum_u1_u3 = u1 + u3) %>%
                        arrange(desc(sum_u1_u3)) %>%
                        head(n=6)
```

6. Look at the extremes of column 2 of U.

```{r}
# Extremes of u2
# Largest
df_movies_key_u %>% arrange(desc(u2)) %>%
                        head(n=5) 
# Smallest
df_movies_key_u %>% arrange(u2) %>%
                        head(n=5)
```

  + comments: The movies with the highest values in both u1 and u3 
have the same characterstics of the two corresponding latent factors.
It is difficult for me to infer some property to be shared by those movies when considering its name and genre.
Similarly, one can sort movies by a latent factor represented by u2. It is hard to tell the differences between the largest and smallest groups with regard to u2.

