---
title: "Homework 1 - readable and efficient R code"
author: "Eunchong Kang"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1 - "function-alize" this code

Read over the code below and perform the following:

* Wrap it into a function `foobar0` which has arguments `x` and `z`
  and which returns the vector `x` at the end of the following code.

```{r}
## 'foobar0' function
set.seed(1)
x <- rnorm(100)
z <- rnorm(100)

foobar0 <- function(x, z) {
        # Sample code
        if (sum(x >= .001) < 1) {
                stop("step 1 requires 1 observation(s) with value >= .001")
        }
        fit <- lm(x ~ z)
        r <- fit$residuals
        x <- sin(r) + .01
        if (sum(x >= .002) < 2) {
                stop("step 2 requires 2 observation(s) with value >= .002")
        }
        fit <- lm(x ~ z)
        r <- fit$residuals
        x <- 2 * sin(r) + .02
        if (sum(x >= .003) < 3) {
                stop("step 3 requires 3 observation(s) with value >= .003")
        }
        fit <- lm(x ~ z)
        r <- fit$residuals
        x <- 3 * sin(r) + .03
        if (sum(x >= .004) < 4) {
                stop("step 4 requires 4 observation(s) with value >= .004")
        }
        fit <- lm(x ~ z)
        r <- fit$residuals
        x <- 4 * sin(r) + .04
        
        # Storing the result in the 'x_foobar0' object in the global environment
        assign("x_foobar0", x, envir = parent.env(environment()))
        
        return(x)
}

# Excuting 'foobar0' function
foobar0(x, z)
```

* Rewrite this into a function `foobar` which is easier to read, by
  reducing repetitive code. E.g. `foobar` might call a function to
  check the input, and another function to perform the three lines of
  computation. 

```{r}
foobar <- function(x, z) {
        
        # Checking input
        cond <- function(i){
                if (sum(x >= 0.001*i) < i){
                        stop(paste("step", i, "requires", i, "observation(s) with value >=", 0.001*i))
                }
        }
        
        # Calculating function
        cal <- function(x, z){
                fit <- lm(x ~z)
                r <- fit$residuals
                assign("x", i*sin(r) + 0.01*i, envir = parent.env(environment()))
        }
        
        # Doing Iteration
        for (i in 1:4){
                cond(i)
                cal(x, z)
        }
        
        # Storing the result in 'x_foobar' object in the global environment
        assign("x_foobar", x, envir = parent.env(environment()))
        
        return(x)
}

# Executing foobar function
foobar(x,z)
```

* Check that the two versions produce the same output using the
  function `all.equal`.
  
```{r}
# Comparing the two results from the functions
all.equal(x_foobar0, x_foobar)
```


# Question 2 - vectorize this code and benchmark

* Take the following function `f0` and rewrite it as a function `f`,
  which is faster and easier to read, by removing the loop of `i` from
  1 to `m`.

```{r}
n <- 30
p <- 50
p2 <- 25
m <- 1000
set.seed(1)
x <- matrix(rnorm(n*p),nrow=n,ncol=p)
b <- matrix(rnorm(m*p),nrow=m,ncol=p)
a <- matrix(rnorm(m*p2),nrow=m,ncol=p2)

# Original f0 function
f0 <- function(x, b, a) {
        out <- numeric(0)
        for (i in seq_len(m)) {
                bb <- b[i, ]
                aa <- a[i, ]
                out <- c(out, sum(x %*% bb) + sum(aa))
        }
        out
}

# Making f function
f <- function(x, b, a){
        colSums(x %*% t(b)) + rowSums(a)
}

# Comparing the results
all.equal(f0(x, b, a), f(x, b, a))
```

* Benchmark `f` and `f0` using `microbenchmark`. How much faster is `f`?

```{r}
library(microbenchmark)
microbenchmark(f0(x, b, a), f(x, b, a))
```
Conclusion: The f function is about five times faster than the f0 function

# Question 3 - build a faster t-test

* Rewrite the following function `getT0` which computes `m`
  two-sample t-tests (equal variance) between two groups as a function
  `getT`, which is faster by using vectorized operations over the `m`
  sets of observations. (There are functions in R packages, such as
  `genefilter::rowttests` which will quickly perform this operation,
  but I want you to write your own function using simple R functions
  like `rowSums`, etc.)

```{r}
m <- 400
n <- 50
little.n <- n/2
set.seed(1)
x <- matrix(rnorm(m*n),nrow=m,ncol=n)
f <- gl(2,little.n)

# Original getT0 function
getT0 <- function(x, f) {
  ts <- sapply(seq_len(m), function(i) t.test(x[i,] ~ f, var.equal=TRUE)$statistic)
  unname(ts)
}

# Making getT function
getT <- function(x){
        # Spliting the data
        x1 <- x[, 1:little.n]
        x2 <- x[, (little.n + 1):ncol(x)]
        
        # Calculating mean difference and pooled sample variance
        mean_diff <- rowMeans(x1) - rowMeans(x2)
        pooled_var <- (rowSums((x1 - rowMeans(x1))^2) + rowSums((x2 - rowMeans(x2))^2))/(n - 2)
        
        # Calculating t-statistic
        mean_diff/sqrt(pooled_var*(1/little.n + 1/little.n))
}

# Comparing the results
all.equal(getT0(x,f), getT(x))
```

* Benchmark `getT` and `getT0`. How much faster is `getT`?

```{r}
microbenchmark(getT0(x,f), getT(x))
```
Conclusion: The getT function is more than 1000 times faster than the getT0 function.
