---
title: "Homework 3 - C++ code from R"
author: "Eunchong Kang"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, warning=FALSE, message=FALSE}
library(usethis)
library(devtools)
library(Rcpp)
library(RcppArmadillo)
library(microbenchmark)
library(MASS)

# load my package
document("../bios735")
load_all("../bios735")
```

# Question 1 - simple Rcpp function

Write a Rcpp function named `one.or.exp` that implements the R code, `ifelse(x < 0.0, 1.0, exp(x))`, and compare its output.

```{r}
x <- -10:10
one.or.exp(x)
all.equal(ifelse(x < 0.0,1.0,exp(x)), one.or.exp(x))
```

# Question 2 - random walk in C++ and vectorized R

Write a random walk function in C++ named `randomWalk2`, and compare its output with
`randomWalk1`.

1. Sample Random walk

```{r}
niter <- 1e4
lambda <- .01
randomWalk1 <- function(niter,lambda) {
  x <- numeric(niter)
  y <- numeric(niter)
  for (i in seq_len(niter)[-1]) {
    x[i] <- x[i-1] + lambda * (2.0 * rbinom(1,1,0.5) - 1.0)
  }
  for (i in seq_len(niter)[-1]) {
    y[i] <- y[i-1] + lambda * (2.0 * rbinom(1,1,0.5) - 1.0)
  }
  list(x=x,y=y)
}
myplot <- function(dat) {
  niter <- length(dat$x)
  plot(0,type="n",xlim=c(-2,2),ylim=c(-2,2),xlab="x",ylab="y")
  cols <- colorRampPalette(c("blue","red","orange"))(100)
  with(dat, segments(x[-niter],y[-niter],x[-1],y[-1],col=rep(cols,each=niter/100)))
}
set.seed(5)
dat1 <- randomWalk1(niter,lambda)
str(dat1)
myplot(dat1)
```

2. my random walk:

```{r}
set.seed(5)
dat2 <- randomWalk2(niter,lambda)
str(dat2)
myplot(dat2)
all.equal(dat2$x, dat1$x)
all.equal(dat2$y, dat1$y)
microbenchmark(randomWalk1(niter,lambda),randomWalk2(niter,lambda),times=10)
```

  + Conclusion: The Rcpp function (`randomWalk2`) is about 30 times faster than 
  the for-loop-in-R function (`randomWalk1`).

<br/>
write a vectorized version of the random walk in R named `randomWalkVectorized`, and
compare its output.


3. Vectorized Random Walk

```{r}
randomWalkVectorized <- function(niter,lambda) {
        x <- c(0, lambda * (2.0 * rbinom(niter-1, 1, 0.5) - 1.0))
        y <- c(0, lambda * (2.0 * rbinom(niter-1, 1, 0.5) - 1.0))
        list(x=cumsum(x), y=cumsum(y))
}
```

```{r}
set.seed(5)
datVec <- randomWalkVectorized(niter,lambda)
str(datVec)
myplot(datVec)
all.equal(datVec$x, dat1$x)
all.equal(datVec$y, dat1$y)
microbenchmark(randomWalk1(niter,lambda),randomWalkVectorized(niter,lambda),times=10)
microbenchmark(randomWalk2(niter,lambda),randomWalkVectorized(niter,lambda),times=10)
```

  + Conclusion: The vectorized funtion (`randomWalkVectorized`) is slightly slower than the Rcpp function (`randomWalk2`), and there is no big difference. However, the Vectorized function is 30 times faster than the for-loop-in-R function (`randomWalk1`)
  

# Question 3 - simple RcppArmadillo function

Write a simple RcppArmadillo function that solves $AX = b$ named `armadilloSolve`.

```{r}
A <- matrix(runif(12),ncol=3)
x <- matrix(runif(3),ncol=1)
b <- A %*% x
xx <- armadilloSolve(A,b)
all.equal(x, xx)
```

# Question 4 - column-wise ridge regression

Write a function `colRidge2` which implements ridge regression.Compare the output of `colRidge2` to `colRidge1`.

```{r}
set.seed(1)
n <- 100
Y <- matrix(rnorm(n*20),nrow=20)
X <- scale(matrix(rnorm(20*2),ncol=2))
lambda <- runif(n,.1,2)

colRidge1 <- function(Y, X, lambda) {
  df <- as.data.frame(X)
  n <- ncol(Y)
  beta <- matrix(nrow=2,ncol=n)
  stopifnot(length(lambda) == n)
  for (j in seq_len(n)) {
    beta[,j] <- coef(lm.ridge(Y[,j] ~ 0 + V1 + V2, data=df, lambda=lambda[j]))
  }
  beta
}
beta1 <- colRidge1(Y, X, lambda)
beta1[,1:5]
beta2 <- colRidge2(Y, X, lambda)
beta2[,1:5]
plot(beta1[1,], beta2[1,])
abline(0,1)
plot(beta1[2,], beta2[2,])
abline(0,1)
all.equal(beta1[1,], beta2[1,])
all.equal(beta1[2,], beta2[2,])
microbenchmark(colRidge1(Y, X, lambda), colRidge2(Y, X, lambda), times=10)
```

  + Conclustion: From the plots and the result of all.eqal function, the results from both functions are very close. The Rcpp function (`ColRidge2`) is much faster than the for-loop-in-R function (`colRidge1`).
