---
title: "HW 7 - Numerical Integration"
author: "Eunchong Kang"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, warning=FALSE}
library(statmod)
library(optimx)
library(numDeriv)
```

# Maximization of poisson GLMM from lecture

Now that we have discussed several approaches for numerical integration, lets now maximize the model given in lecture.  You may choose any maximization approach, as well as any numerical integration procedure from lecture, to obtain the MLE's for $\boldsymbol{\beta}$ and $\sigma_{\gamma}^2$.  

Hint: You should evaluate a number of intervals/nodes, decreasing the convergence threshold, etc and evaluate its impact before reporting your final result. We have shown how to perform, for example, AGQ and IS to obtain the likelihood pertaining to the first subject from class. 

```{r}
setwd("D:/Spring_2022_D/BIOS735/hw/hw7")
alz = read.table("alzheimers.dat", header = T)
xi <- cbind(rep(1,5), 1:5)
y1 <- alz$words[alz$subject == 1]
beta = c(1.804, 0.165)
s2gamma = 0.000225

## 1. Integration related helper functions
# 1-1. Create integrand function for ith subject
# This needs to calculate muhat and sigma2hat for aGQ
inner <- function(gammai, yi, beta, s2gamma, log = F){
        val <- rep(NA, length(gammai))
        
        for (i in 1:length(val)){
                lambda <- exp(xi %*% beta + gammai[i])
                val0 <- dpois(x = yi, lambda = lambda)
                val[i] <- prod(val0) * dnorm(x = gammai[i],
                                             mean = 0,
                                             sd = sqrt(s2gamma))
        }
        
        if (log == F){
                return(val)
        } else{
                return(log(val))
        }
}
# Confirm
inner(0.2, y1, beta, s2gamma)

# 1-2. Create a function calculating the posterior mode of the integrand (=muhat)
muhat <- function(yi, beta, s2gamma) {
        pm <- suppressWarnings(
                optimx(
                        par = 0,
                        fn = function(x, yi, beta, s2gamma) {
                                inner(
                                        gammai = x,
                                        yi = yi,
                                        beta = beta,
                                        s2gamma = s2gamma
                                )
                        },
                        yi = yi,
                        beta = beta,
                        s2gamma = s2gamma,
                        method = "Nelder-Mead",
                        control = list(maximize = T)
                )
        )
        return(pm$p1)
}
# Confirm
muhat(y1, beta, s2gamma)

# 1-3. Create a function calculating a sigma2hat
sigma2hat <- function(muhat, yi, beta, s2gamma, inner) {
        hess_pm <- hessian(
                func = inner,
                x = muhat,
                yi = yi,
                beta = beta,
                s2gamma = s2gamma,
                log = T
        )
        
        return(-1 / hess_pm)
}
# 사용법 결과=matrix
sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner)
sqrt(sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner))

# 1-4 Create h(x) function 
# Integrad factors into h(x)*pi(gammai|muhat,sigma2hat)
h_t <- function(t, yi, beta, s2gamma, muhat, sigma2hat){
        val <- rep(NA, length(t))
        
        for (i in 1:length(val)){
                lambda <- exp(xi %*% beta + t[i])
                val0 <- dpois(x = yi, lambda = lambda)
                val[i] <- prod(val0) * 
                        dnorm(x = t[i], mean = 0, sd = sqrt(s2gamma)) / 
                        dnorm(x = t[i], mean = muhat, sd = sqrt(sigma2hat))
        }
        return(val)
}

# Confirm
h_t(t, y1, beta, s2gamma, muhat(y1, beta, s2gamma), sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner))

M <- 5
gh <- gauss.quad(n = M, kind = "hermite")
nodes_over_gamma1 <- gh$nodes

t <- muhat(y1, beta, s2gamma) + sqrt(2*as.numeric(sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner))) * nodes_over_gamma1

w <- gh$weights /sqrt(pi)

val <- sum(w * h_t(t, y1, beta, s2gamma, muhat(y1, beta, s2gamma), sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner)))

val


## 2. Integrate 
# This function calculates Log-likelihood given beta and s2gamma
integr <- function(theta, M) {
        # Split theta into beta and s2gama
        beta <- theta[1:2]
        s2gamma <- theta[3]
        
        # Finds nodes and weights
        gh <- gauss.quad(n = M, kind = "hermite")
        nodes_over_gamma <- gh$nodes
        
        # Create NA vector
        val <- rep(NA, max(alz$subject))
        
        # Looping
        for (i in 1:max(alz$subject)) {
                # Subset data by subject
                yi <- alz$words[alz$subject == i]
                
                # Calculate muhat and sigma2hat by subject
                muhat_yi <- muhat(yi, beta, s2gamma)
                sigma2hat_yi <- sigma2hat(muhat_yi, yi, beta, s2gamma, inner)
                
                # Calculate new nodes and weights
                t <- muhat_yi + sqrt(2 * as.numeric(sigma2hat_yi)) * nodes_over_gamma
                w <- gh$weights / sqrt(pi)
                
                # Approximate the integral by subject
                val[i] <- sum(
                        w * h_t(
                                t = t,
                                yi = yi,
                                beta = beta,
                                s2gamma = s2gamma,
                                muhat = muhat_yi,
                                sigma2hat = sigma2hat_yi
                        )
                )
                
        }
        
        # Calculate Log-likelihood
        return(sum(log(val)))
}
# Confirm
integr(c(beta, s2gamma), M=5)


## 3. Optimize using Nelder-Mead
fit <- optimx(
        par = c(beta, s2gamma),
        fn = function(x, M){integr(theta=x, M)},
        M = 1,
        method = "Nelder-Mead",
        control = list(maximize = T)
)

```

# Plot

Now, plot the fitted line from the fitted GLMM on the spaghetti plot from lecture

```{r}
## solution
## end solution
```