---
title: "HW 7 - Numerical Integration"
author: "Eunchong Kang"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, warning=FALSE}
library(statmod)
library(optimx)
library(numDeriv)
```

# Maximization of poisson GLMM from lecture

Now that we have discussed several approaches for numerical integration, lets now maximize the model given in lecture.  You may choose any maximization approach, as well as any numerical integration procedure from lecture, to obtain the MLE's for $\boldsymbol{\beta}$ and $\sigma_{\gamma}^2$.  

Hint: You should evaluate a number of intervals/nodes, decreasing the convergence threshold, etc and evaluate its impact before reporting your final result. We have shown how to perform, for example, AGQ and IS to obtain the likelihood pertaining to the first subject from class. 

```{r}
## Integration related helper functions
# 1. Creating integrand function
inner <- function(gammai, yi, xi, beta, s2gamma, log = F){
        val <- rep(NA, length(gammai))
        
        for (i in 1:length(val)){
                lambda <- exp(xi %*% beta + gammai[i])
                val0 <- dpois(x = yi, lambda = lambda)
                val[i] <- prod(val0) * dnorm(x = gammai[i],
                                             mean = 0,
                                             sd = sqrt(s2gamma))
        }
        
        if (log == F){
                return(val)
        } else{
                return(log(val))
        }
}
# 사용법
inner(c(1,0.003), y1, xi, beta, s2gamma)



# 1-2. Create a function calculating the posterior mode of the integrand (mu_hat)
muhat <- function(yi, beta, s2gamma) {
        pm <- suppressWarnings(
                optimx(
                        par = 0,
                        fn = function(x, yi, xi, beta, s2gamma) {
                                inner(
                                        gammai = x,
                                        xi = xi,
                                        yi = yi,
                                        beta = beta,
                                        s2gamma = s2gamma
                                )
                        },
                        yi = yi,
                        xi = xi,
                        beta = beta,
                        s2gamma = s2gamma,
                        method = "Nelder-Mead",
                        control = list(maximize = T)
                )
        )
        return(pm$p1)
}
# ex
muhat(y1, beta, s2gamma)

         yyyy<-       optimx(
                        par = 0,
                        fn = function(x, yi, xi, beta, s2gamma) {
                                inner(
                                        gammai = x,
                                        xi = xi,
                                        yi = yi,
                                        beta = beta,
                                        s2gamma = s2gamma
                                )
                        },
                        yi = y1,
                        xi = xi,
                        beta = beta,
                        s2gamma = s2gamma,
                        method = "Nelder-Mead",
                        control = list(maximize = T)
                )

y1 <- alz$words[alz$subject == 1]
















# 1-3. Find a sigma_hat sqaure
hess_pm <- hessian(
        func = inner,
        x = pm(y1, beta, s2gamma),
        yi = y1,
        beta = beta,
        s2gamma = s2gamma,
        log = T
)
#sqrt(-1 / hess_pm) is the sigma_hat square
sigma2hat <- function(muhat, yi, beta, s2gamma, inner) {
        hess_pm <- hessian(
                func = inner,
                x = muhat,
                xi = xi,
                yi = yi,
                beta = beta,
                s2gamma = s2gamma,
                log = T
        )
        
        return(-1 / hess_pm)
}
# 사용법 결과=matrix
sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner)
sqrt(sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner))


# Create h(x) function
h_t <- function(t, yi, xi, beta, s2gamma, muhat, sigma2hat){
        val <- rep(NA, length(t))
        
        for (i in 1:length(val)){
                lambda <- exp(xi %*% beta + t[i])
                val0 <- dpois(x = yi, lambda = lambda)
                val[i] <- prod(val0) * 
                        dnorm(x = t[i], mean = 0, sd = sqrt(s2gamma)) / 
                        dnorm(x = t[i], mean = muhat, sd = sqrt(sigma2hat))
        }
        return(val)
}

# 쓰는 법
h_t(t, y1, xi, beta, s2gamma, muhat(y1, beta, s2gamma), sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner))



M <- 5
gh <- gauss.quad(n = M, kind = "hermite")
nodes_over_gamma1 <- gh$nodes

t <- muhat(y1, beta, s2gamma) + sqrt(2*as.numeric(sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner))) * nodes_over_gamma1

w <- gh$weights /sqrt(pi)

val <- sum(w * h_t(
        t = t,
        yi = y1,
        beta = beta,
        s2gamma = s2gamma,
        pm = pm$p1,
        hess_pm = hess_pm
))


#이게 i= 1일때 적분 값. 총 22번 돌리고 곱해야 함

y1 = alz$words[alz$subject == 1]

max(alz$subject)


sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner)
h_t(t, y1, xi = cbind(rep(1,5), 1:5), beta, s2gamma, muhat(y1, beta, s2gamma), sigma2hat(muhat(y1, beta, s2gamma), y1, beta, s2gamma, inner))
h_t <- function(t, yi, xi = cbind(rep(1,5), 1:5), beta, s2gamma, muhat, sigma2hat)

        

# Finds nodes and weights
gh <- gauss.quad(n = M, kind = "hermite")
nodes_over_gamma1 <- gh$nodes

# Create NA vector
val <- rep(NA, max(alz$subject))

# Looping
for (i in 1:max(alz$subject)){

        yi <- alz$words[alz$subject == i]
        
        muhat_yi <- muhat(yi, beta, s2gamma)
        sigma2hat_yi <- sigma2hat(muhat_yi, yi, beta, s2gamma, inner)
        
        #나중에 gamma1 > gamma로 바꾸면 될듯
        t <- muhat_yi + sqrt(2 * as.numeric(sigma2hat_yi)) * nodes_over_gamma1
        w <- gh$weights /sqrt(pi)
        
        # Approximate the integral
        val[i] <- sum(w * h_t(
                t = t,
                yi = yi,
                xi = xi,
                beta = beta,
                s2gamma = s2gamma,
                muhat = muhat_yi,
                sigma2hat = sigma2hat_yi
        ))
        
}

prod(val)



beta
s2gamma

#integr <- function(theta, M, dt = alz, inner=inner, xi=cbind(rep(1,5), 1:5)) {

integr <- function(theta, M) {
        ##추가
        beta <- theta[1:2]
        s2gamma <- theta[3]
        
        # Finds nodes and weights
        gh <- gauss.quad(n = M, kind = "hermite")
        nodes_over_gamma1 <- gh$nodes
        
        # Create NA vector
        val <- rep(NA, max(alz$subject))
        
        # Looping
        for (i in 1:max(alz$subject)) {
                yi <- alz$words[alz$subject == i]
                
                muhat_yi <- muhat(yi, beta, s2gamma)
                sigma2hat_yi <- sigma2hat(muhat_yi, yi, beta, s2gamma, inner)
                
                #나중에 gamma1 > gamma로 바꾸면 될듯
                t <- muhat_yi + sqrt(2 * as.numeric(sigma2hat_yi)) * nodes_over_gamma1
                w <- gh$weights / sqrt(pi)
                
                # Approximate the integral
                val[i] <- sum(
                        w * h_t(
                                t = t,
                                yi = yi,
                                xi = xi,
                                beta = beta,
                                s2gamma = s2gamma,
                                muhat = muhat_yi,
                                sigma2hat = sigma2hat_yi
                        )
                )
                
        }
        sum(log(val))
        #return(val)
        
        #if (is.nan(prod(val)) == T){
        #       return(4.629948e-265)
        #} else{return(prod(val))}

}

# 4.881176e-158
x <- c(beta, s2gamma)

integr(beta, s2gamma, M = 5, dt = alz, inner=inner, xi=cbind(rep(1,5), 1:5))

#-362.2231
integr(c(beta, s2gamma), M=5)
integr(c(beta, s2gamma))
log(4.881176e-158)

integr(x, M = 5, dt = alz, inner=inner, xi=cbind(rep(1,5), 1:5))

integr(theta=c(1.804, 0.165, 0.000225), M = 5, dt = alz, inner=inner, xi=cbind(rep(1,5), 1:5))

integr(theta=c(1.204, 1.5, 0.1), M = 5)
beta
s2gamma



# function(x, M, dt, inner, xi){
#                 integr(x, M = M, dt = dt, inner = inner, xi = xi)
#         },

beta
s2gamma


fit <- optimx(
        par = c(beta, s2gamma),
        fn = function(x, M){integr(theta=x, M)},
        M = 1,
        method = "Nelder-Mead",
        control = list(
                #trace = 0, # higher number print more detailed output
                maximize = T
                # parscale = vector of scaling values to correct for scale differencs in parameters.  Should be the same length as the num of parmaters
        )
)

## results p1=1.804, p2=0.3454, p3=0.000225 : minimize
## results p1=1.804, p2=0.165, p3=0.180625 : maximize

integr(theta=c(1.804, 0.05, 0.1))
integr(theta=c(1.804, 0.165, 0.180625))
integr(theta=c(1.804, 0.165, 0.30625))
integr(theta=c(1.4, 0.165, 0.9))
#1.820155 0.1657221 0.2182965
integr(theta=c(1.820155, 0.1657221, 0.2182965))


fit <- optimx(
        par = c(1.804, 0.165, 0.000225),
        fn = function(x){integr(theta=x)},
        method = "Nelder-Mead",
        control = list(
                #trace = 0, # higher number print more detailed output
                maximize = T
                # parscale = vector of scaling values to correct for scale differencs in parameters.  Should be the same length as the num of parmaters
        )
)










integr(c(beta, s2gamma), M=5)

str(alz)

c(beta, s2gamma)[1:2]

c(beta, s2gamma)[3]

beta














val <- sum(w * h_t(
        t = t,
        yi = y1,
        beta = beta,
        s2gamma = s2gamma,
        pm = pm$p1,
        hess_pm = hess_pm
))

?optimx


## Solution: place relevant helper functions pertaining to integration here 
## End Solution




## Solution: place relevant helper functions pertaining to maximization here (likelihood etc)
## End Solution







## Solution: place primary code for maximization here, calling functions in the above two sections




## Remember to print your primary results and use the following starting values
beta = c(1.804, 0.165)
s2gamma = 0.000225
xi <- cbind(rep(1,5), 1:5)

setwd("D:/Spring_2022_D/BIOS735/hw/hw7")
alz = read.table("alzheimers.dat", header = T)
## End Solution
```

# Plot

Now, plot the fitted line from the fitted GLMM on the spaghetti plot from lecture

```{r}
## solution
## end solution
```